Towards a Verified Tableau Prover for a Quantifier-Free Fragment of Set Theory

### 2023-03-03 paper/CADE/paper.tex
A few amendments

\subsection{Contributions}
We present a formalisation in Isabelle/HOL of a tableau calculus for \MLSS{} due to \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}\cite[Chapter 14]{set_theory}.
We prove soundness and completeness of the calculus and give an abstract specification of a decision procedure that applies the rules of the calculus exhaustively.
To obtain total correctness of the procedure, we prove its termination.
We also formalise a refinement to an executable specification from which Isabelle can generate code.
The formalisation initially follows the paper but gives a more thorough account of some important details:
\begin{itemize}
  \item We deliver the omitted proof of Lemma~2 in the paper~\cite{new_fast_tableau}, which is a key building block for the completeness proof of the calculus.
  \item The formal proof of completeness lead to the discovery that the calculus was missing a rule for eliminating double negation. 
  \item We derive an explicit upper bound for the number of formulae in a branch of the tableau.
\end{itemize}

@@

We abstractly specify the behaviour of the decision procedure, leaving an executable specification from which code can be generated to future work.
The implementation uses a couple of features of Isabelle/HOL's function package:
instead of defining the function via pattern matching, we specify the equations of the function as conditional rewrite rules.
This requires us to prove that the assumptions of the equations are non-overlapping, which is done by automation.
The other concern is that Isabelle/HOL requires functions to be total, so a recursive function needs to terminate in order for it to be well-defined;
nevertheless, the termination proof is separated from the definition of the function for modularity.
The function package maintains the soundness of the definition by introducing a so-called domain predicate \lstinline!mlss_proc_branch_dom! which characterises the arguments for which the function terminates.
Each equation of the function is guarded by an assumption that the predicate holds for the argument.
In \autoref{sec:correct}, we will show that the domain predicate holds for the context that \lstinline!mlss_proc_branch! is called in.
Before we go into more detail on how the termination is proved, we discuss its definition as shown below.

@@

\section{Conclusion and Future Work}
By closely following a paper by \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}, we developed a formalisation of a tableau calculus for a quantifier-free fragment of set theory called \MLSS{}.
The formalisation includes an abstract description of a decision procedure that builds on the calculus.
To make the decision procedure compatible with Isabelle/HOL, we extended the calculus with a light-weight type system while maintaining completeness.

The refinement of the abstract description to an executable specification from which code can be exported is left for future work.
This executable specification should generate certificates that can be replayed through Isabelle's inference kernel in order to facilitate the integration of the procedure into Isabelle.

@@

Concerning the second inequality, first observe that each literal contains an atom, implying that the set of literals is at most twice as large as the set of atoms in a branch. 
In addition, an atom is formed by combining two subterms occurring in the branch with either \lstinline!$\inS$! or \lstinline!$\eqS$!;
hence, we have
\begin{lstlisting}
  &\phantom{$\leq$}& |{$\psi$ $\in$ set b. is_literal $\psi$}|
  $\leq$ 2 * |$\bigcup$(atoms ` set b)| $\leq$ 2 * 2 * |subterms b|$^\text{2}$.
\end{lstlisting}
Since no expansion rule introduces new subterms, we know the cardinality of the subterms in a branch \lstinline!b! is limited by the subterms of \lstinline!last b! plus the number of witnesses added to the branch.
Note that the last branching rule requires \lstinline!s $\neqS$ t! to be in \lstinline!b! for some \lstinline!s,t $\in$ subterms (last b)!.
The subsumption condition then prohibits that the rule can be applied again for the combination of \lstinline!s! and \lstinline!t!, thus yielding an upper bound of \lstinline!|subterms (last b)|$^\text{2}$! for the witnesses.
This allows us to finish the calculation that justifies the second inequality:
\begin{lstlisting}
  &\phantom{$\leq$}& 4 * |subterms b|$^\text{2}$
  $\leq$ 4 * |subterms (last b)| + |wits b|$^\text{2}$
  $\leq$ 4 * |subterms (last b)| + (|subterms (last b)|$^\text{2}$)$^\text{2}$
  $\leq$ 4 * 4 * |subterms (last b)|$^\text{4}$.
\end{lstlisting}
All in all, we get the following lemma.
\begin{lstlisting}[label={lst:card_branch}]
lemma card_wf_branch_ub:
  assumes wf_branch b
  shows |set b|
      $\leq$ 2 * |subfms (last b)| + 16 * |subterms (last b)|$^\text{4}$
\end{lstlisting}

>>>>

\subsection{Contributions}
We present a formalisation in Isabelle/HOL of a tableau calculus for \MLSS{} due to \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}\cite[Chapter 14]{set_theory}.
We prove soundness and completeness of the calculus and give an abstract specification of a decision procedure that applies the rules of the calculus exhaustively.
To obtain total correctness of the procedure, we prove its termination.
In order to obtain an executable procedure, we also give a naive refinement to a more concrete specification from which code can be generated.
The formalisation initially follows the paper but gives a more thorough account of some important details:
\begin{itemize}
  \item We deliver the omitted proof of Lemma~2 in the paper~\cite{new_fast_tableau}, which is a key building block for the completeness proof of the calculus.
  \item The formal proof of completeness lead to the discovery that the calculus was missing a rule for eliminating double negation. 
  \item We derive an explicit upper bound for the number of formulae in a branch of the tableau.
\end{itemize}

@@

We only discuss the abstract specification here and refer the reader to the formalisation for the executable specification.
The implementation uses a couple of features of Isabelle/HOL's function package:
instead of defining the function via pattern matching, we specify the equations of the function as conditional rewrite rules.
This requires us to prove that the assumptions of the equations are non-overlapping, which is done by automation.
The other concern is that Isabelle/HOL requires functions to be total, so a recursive function needs to terminate in order for it to be well-defined;
nevertheless, the termination proof is separated from the definition of the function for modularity.
The function package maintains the soundness of the definition by introducing a so-called domain predicate \lstinline!mlss_proc_branch_dom! which characterises the arguments for which the function terminates.
Each equation of the function is guarded by an assumption that the predicate holds for the argument.
In \autoref{sec:correct}, we will show that the domain predicate holds for the context that \lstinline!mlss_proc_branch! is called in.
Before we go into more detail on how the termination is proved, we discuss its definition as shown below.

@@

\section{Conclusion and Future Work}
By closely following a paper by \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}, we developed a formalisation of a tableau calculus for a quantifier-free fragment of set theory called \MLSS{}.
The formalisation includes an abstract description of a decision procedure that builds on the calculus.
To make the decision procedure compatible with Isabelle/HOL, we extended the calculus with a light-weight type system while maintaining completeness.
We also refined the abstract specification to an executable specification from which code can be generated.

In future work, we plan to implement an efficient executable specification in the style of a worklist algorithm. 
This specification should also generate certificates that can be replayed through Isabelle's inference kernel in order to facilitate the integration of the procedure into Isabelle.

@@

Concerning the second inequality, first observe that each literal contains an atom, implying that the set of literals is at most twice as large as the set of atoms in a branch. 
In addition, an atom is formed by combining two subterms occurring in the branch with either \lstinline!$\inS$! or \lstinline!$\eqS$!;
hence, we have
\begin{lstlisting}
  &\phantom{$\leq$}& |{$\psi$ $\in$ set b. is_literal $\psi$}|
  $\leq$ 2 * |$\bigcup$(atoms ` set b)| $\leq$ 2 * 2 * |subterms b|$^\text{2}$.
\end{lstlisting}
Since no expansion rule introduces new subterms, we know the cardinality of the subterms in a branch \lstinline!b! is limited by the subterms of \lstinline!last b! plus the number of witnesses added to the branch.
Note that the last branching rule requires \lstinline!s $\neqS$ t! to be in \lstinline!b! for some \lstinline!s,t $\in$ subterms (last b)!.
The subsumption condition then prohibits that the rule can be applied again for the combination of \lstinline!s! and \lstinline!t!, thus yielding an upper bound of \lstinline!|subterms (last b)|$^\text{2}$! for the witnesses.
This allows us to finish the calculation that justifies the second inequality:
\begin{lstlisting}
  &\phantom{$\leq$}& 4 * |subterms b|$^\text{2}$
  $\leq$ 4 * |subterms (last b)| + |wits b|$^\text{2}$
  $\leq$ 4 * |subterms (last b)| + (|subterms (last b)|$^\text{2}$)$^\text{2}$
  $\leq$ 4 * 4 * |subterms (last b)|$^\text{4}$.
\end{lstlisting}
All in all, we get the following lemma.
\begin{lstlisting}
lemma card_wf_branch_ub:
  assumes wf_branch b
  shows |set b|
      $\leq$ 2 * |subfms (last b)| + 16 * |subterms (last b)|$^\text{4}$
\end{lstlisting}
